---
title: "Supervised Learning I - K-Nearest Neighbors and Resampling Methods"
author: "Patrick PUN Chi Seng (NTU Sg)"
output:
  html_document:
    df_print: paged
  pdf_document: default
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyfoot[C]{\thepage}
- \fancypagestyle{plain}{\pagestyle{fancy}}
- \fancyhead[L]{Machine Learning in Finance}
- \fancyhead[R]{R Lab - KNN and Resampling}
---


```{r global_options, include=FALSE}
library(knitr)
knitr::opts_chunk$set(eval=TRUE)
```

## References

Chapters 4.7.1, 4.7.6, 5.3 [ISLR2] An Introduction to Statistical Learning - with Applications in R (2nd Edition). Free access to download the book: https://www.statlearning.com/

To see the help file of a function `funcname`, type `?funcname`.

## *K* Nearest Neighbors (KNN)

We will begin by examining some numerical and graphical summaries of the `Smarket` data, which is part of the `ISLR2` library. This data set consists of percentage returns for the S\&P 500 stock index over $1,250$ days, from the beginning of 2001 until the end of 2005. For each date, we have recorded the percentage returns for each of the five previous trading days, `lagone` through `lagfive`. We have also recorded `volume` (the number of shares traded on the previous day, in billions), `Today` (the percentage return on the date in question)  and `direction` (whether the market was `Up` or `Down` on this date). Our goal is to predict `direction` (a qualitative response) using the other features.

```{r chunk1}
library(ISLR2)
names(Smarket)
dim(Smarket)
summary(Smarket)
pairs(Smarket)
```

The `cor()` function produces a matrix that contains all of the pairwise correlations among the predictors in a data set. The first command below gives an error message because the `direction` variable is qualitative.

```{r chunk2, error=TRUE}
cor(Smarket)
cor(Smarket[, -9])
```

As one would expect, the correlations between the lag variables and today's returns are close to zero. In other words, there appears to be little correlation between today's returns and previous days' returns. The only substantial correlation is between `Year` and `volume`. By plotting the data, which is ordered chronologically, we see that `volume` is increasing over time. In other words, the average number of shares traded daily increased from 2001 to 2005.

```{r chunk3}
attach(Smarket)
plot(Volume)
```

We will now perform KNN using the `knn()` function, which is part of the `class` library. `knn()` forms predictions using a single command. The function requires four inputs.

* A matrix containing the predictors associated with the training data, labeled `train.X` below.
* A matrix containing the predictors associated with the data for which we wish to make predictions, labeled `test.X` below.
* A vector containing the class labels for the training observations, labeled `train.Direction` below.
* A value for $K$, the number of nearest neighbors to be used by the classifier.

We use the `cbind()` function, short for *column bind*, to bind the `Lag1` and `Lag2` variables together into two matrices, one for the training set and the other for the test set.

```{r chunk26}
library(class)
train <- (Year < 2005)
data <- cbind(Lag1, Lag2)
train.X <- data[train,]
test.X <- data[!train,]
train.Direction <- Direction[train]
test.Direction <- Direction[!train]
```

Now the `knn()` function can be used to predict the market's movement for the dates in 2005. We set a random seed before we apply `knn()` because if several observations are tied as nearest neighbors, then `R` will randomly break the tie. Therefore, a seed must be set in order to ensure reproducibility of results.

```{r chunk27}
set.seed(1)
knn.pred <- knn(train.X, test.X, train.Direction, k = 1)
table(knn.pred, test.Direction)
mean(knn.pred == test.Direction)
```

The results using $K=1$ are not very good,  since only $50$ \% of the observations are correctly predicted. Of course, it may be that $K=1$ results in  an overly flexible fit to the data. Below, we repeat the analysis using $K=3$.

```{r}
set.seed(1)
knn.pred <- knn(train = train.X, test = test.X, cl = train.Direction, k = 3)
table(knn.pred, test.Direction)
mean(knn.pred == test.Direction)
# Compare the performance of K=1 and K=3.
```

The results have improved slightly. But increasing $K$ further turns out to provide no further improvements.

KNN does not perform well on the `Smarket` data but it does often provide impressive results. As an example we will apply the KNN approach to the `Insurance` data set, which is part of the `ISLR2` library. This data set includes $85$ predictors that measure demographic characteristics for 5,822 individuals. The response variable is `Purchase`, which indicates whether or not a given individual purchases a caravan insurance policy. In this data set, only $6$ \% of people purchased caravan  insurance.

```{r chunk29}
dim(Caravan)
attach(Caravan)
summary(Purchase)
348 / 5822
```

Because the KNN classifier predicts the class of a given test observation by identifying the observations that are nearest to it, the scale of the variables matters. Variables that are on a large scale will have a much larger effect on the *distance* between the observations, and hence on the KNN classifier, than variables that are on a small scale. For instance, imagine a data set that contains two variables, `salary` and `age` (measured in dollars and years, respectively). As far as KNN is concerned, a difference of $\$1,000$ in salary is enormous compared to a difference of $50$ years in age. Consequently, `salary` will drive the KNN classification results, and `age` will have almost no effect. This is contrary to our intuition  that a salary difference of $\$1,000$ is quite small compared to an age difference of $50$ years. Furthermore, the importance of scale to the KNN classifier leads to another issue: if we measured `salary` in Japanese yen, or if we measured `age` in minutes, then we'd get quite different classification results from what we get if these two variables are measured in dollars and years.

A good way to handle this problem is to *standardize* the data so that all variables are given a mean of  zero and  a standard deviation of one. Then all variables will be on a comparable scale. The `scale()` function does just this. In standardizing the data, we exclude column $86$, because that is the qualitative `Purchase` variable.

Note that we cannot standardize the whole dataset and then split it into training and test data, because it leads to test data leakage to training data. We now first split the observations into a test set, containing the first 1,000 observations, and a training set, containing the remaining observations. Then we standardize the data with training data features' means and standard deviations. We fit a KNN model on the training data using $K=1$, and evaluate its performance on the test data.

```{r chunk31}
test <- 1:1000
train.X <- Caravan[-test, -86]
test.X <- Caravan[test,-86]
train.X <- scale(train.X)
train_means <- attr(train.X,"scaled:center")
train_sds <- attr(train.X, "scaled:scale")
test.X <- scale(test.X, train_means, train_sds)
train.Y <- Purchase[-test]
test.Y <- Purchase[test]

set.seed(1)
knn.pred <- knn(train.X, test.X, train.Y, k = 1)
mean(test.Y != knn.pred)
mean(test.Y != "No")
```

The vector `test` is numeric, with values from  $1$ through $1,000$. Typing `Caravan[test, -86]` yields the submatrix of the data containing the observations whose indices range from $1$ to $1,000$, whereas typing `Caravan[-test, -86]` yields the submatrix containing the observations whose indices do *not* range from $1$ to $1,000$. The KNN error rate on the 1,000 test observations is just under $12$ \%. At first glance, this may appear to be fairly good. However, since only $6$ \% of customers purchased insurance, we could get the error rate down to $6$ \% by always predicting `No` regardless of the values of the predictors!

Suppose that there is some non-trivial cost to trying to sell insurance to a given individual. For instance, perhaps a salesperson must  visit each potential customer. If the company tries to sell insurance to a random selection of customers, then the success rate will be only $6$ \%, which may be far too low given the costs involved. Instead, the company would like to try to sell insurance only to customers who are likely to buy it. So the  overall error rate is not of interest. Instead,  the fraction of individuals that are correctly predicted to buy insurance is of interest.

It turns out that KNN with $K=1$ does far better than random guessing among the customers that are predicted to buy insurance. Among  $77$ such customers, $9$, or $11.7$ \%, actually do purchase insurance. This is  double the rate that one would obtain from random guessing.

```{r chunk32}
table(knn.pred, test.Y)
9 / (67 + 9)
```

Using $K=3$, the success rate increases to $23.1$ \%, and with $K=5$ the rate is also $23.1$ \%. This is over four times the rate that results from random guessing. It appears that KNN is finding some real patterns in a difficult data set!

```{r chunk33}
knn.pred <- knn(train.X, test.X, train.Y, k = 3)
table(knn.pred, test.Y)
6 / 26
knn.pred <- knn(train.X, test.X, train.Y, k = 5)
table(knn.pred, test.Y)
3 / 13
```

However, while this strategy is cost-effective, it is worth noting that only 13 customers are predicted to purchase insurance using KNN with $K=5$. In practice, the insurance company may wish to expend resources on convincing more than just 13 potential customers to buy insurance.

## Resampling Methods

In practice, we often need to evaluate model performance and choose the best model among candidates. Resampling methods provide tools for estimating the test error, which helps us avoid overfitting and improves generalization. We illustrate these methods using the `Smarket` dataset with KNN classification, and later with bootstrap and (over/under)sampling examples.

```{r}
train <- (Year < 2005)
data <- cbind(Lag1, Lag2)
train.X <- data[train,]
test.X <- data[!train,]
train.Direction <- Direction[train]
test.Direction <- Direction[!train]
```

### Validation-set Approach

The validation-set approach involves splitting the available data into two parts: a training set and a validation (test) set. The model is trained on the training set and evaluated on the validation set.

In the following code, we randomly select 70\% of the data for training and use the remaining 30\% for testing. We then compute test error rates for different values of k in KNN.
```{r}
set.seed(1)
ntrain=nrow(train.X)
pdtrainind <- sample(ntrain, 0.7*ntrain)
pdtrain.X <- train.X[pdtrainind,]
pdtest.X <- train.X[-pdtrainind,]
pdtrain.Direction <- train.Direction[pdtrainind]
pdtest.Direction <- train.Direction[-pdtrainind]
ErrTe=0; Kmax=50
for(k in 1:Kmax){
  knn.pred <- knn(train = pdtrain.X, test = pdtest.X, cl = pdtrain.Direction, k = k)
  ErrTe[k]=mean(knn.pred != pdtest.Direction)
}
plot(ErrTe,type="l")
which.min(ErrTe)
```

The plot shows how the error rate changes with k. The which.min(ErrTe) function returns the value of k with the lowest validation error. Repeating the split with a different random seed may give a different “optimal” k, showing that the validation-set approach can be unstable.

```{r}
set.seed(2)
ntrain=nrow(train.X)
pdtrainind <- sample(ntrain, 0.7*ntrain)
pdtrain.X <- train.X[pdtrainind,]
pdtest.X <- train.X[-pdtrainind,]
pdtrain.Direction <- train.Direction[pdtrainind]
pdtest.Direction <- train.Direction[-pdtrainind]
ErrTe=0; Kmax=50
for(k in 1:Kmax){
  knn.pred <- knn(train = pdtrain.X, test = pdtest.X, cl = pdtrain.Direction, k = k)
  ErrTe[k]=mean(knn.pred != pdtest.Direction)
}
plot(ErrTe,type="l")
which.min(ErrTe)
```

### F-fold cross Validation
To obtain a more reliable estimate, we can use k-fold cross validation. The data is split into F roughly equal parts (“folds”). Each fold is used once as a validation set, while the other folds form the training set. The test errors are averaged across folds.

Here, we use 10-fold cross validation:
```{r}
F <- 10
set.seed(1)
folds <- cut(seq(1,ntrain), breaks = F, labels = FALSE)
folds <- folds[sample(ntrain)]
folds
```

We then compute the error rates for k=1,...,50 and average across folds:
```{r}
Kmax=50
ErrCV <- matrix(0, nrow=F, ncol=Kmax)
for (f in 1:F) {
  
  # Pseudo train-test split
  pdtrainind <- which(folds != f)
  pdtrain.X <- train.X[pdtrainind,]
  pdtest.X <- train.X[-pdtrainind,]
  pdtrain.Direction <- train.Direction[pdtrainind]
  pdtest.Direction <- train.Direction[-pdtrainind]
  
  for(k in 1:Kmax){
    # Model fitting
    knn.pred <- knn(train = pdtrain.X, test = pdtest.X, cl = pdtrain.Direction, k = k)
    
    # Error
    ErrCV[f,k]=mean(knn.pred != pdtest.Direction)
  }
}
CV <- apply(ErrCV, 2, mean)
plot(CV,type="l")
which.min(CV)
```
The resulting plot typically shows a smoother curve than the validation-set approach. The `which.min(CV)` gives the best choice of k.

### Bootstrap
Bootstrap is another powerful resampling technique. Instead of splitting data into folds, we generate many bootstrap datasets by sampling with replacement from the training set. Each bootstrap dataset is the same size as the original, but contains duplicate observations. This allows us to estimate variability of a statistic or model parameter.

Create one bootstrap dataset:
```{r}
BSind<-sample(ntrain,replace=TRUE)
BSdata<-train.X[BSind,]
```

Generate multiple bootstrap datasets using a `for-loop`:
```{r}
B=100
for(b in 1:B){
  BSind<-sample(ntrain,replace=TRUE)
  BSdata<-train.X[BSind,]
  # ...perform analyses...
  # ...store the results indexed by b...
}
```

Bootstrap is particularly useful for estimating the standard error of model parameters when no analytical formula is available.

### Oversampling and Undersampling

In imbalanced classification problems, where one class is much rarer than the other, resampling techniques can create a more balanced dataset.

We illustrate this with a subset of the iris dataset containing only setosa (majority) and versicolor (minority).
```{r}
irisdata<-iris[1:75,]
irisdata$Species <- factor(irisdata$Species)
n<-nrow(irisdata)
majorind=(1:n)[irisdata[,5]=="setosa"]
minorind=(1:n)[irisdata[,5]=="versicolor"]
majorn=length(majorind)
minorn=length(minorind)
```

*Oversampling* duplicates minority-class samples until class sizes are equal.
```{r}
OSind=sample(minorind,majorn-minorn,replace=TRUE)
OSdata<-rbind(irisdata,irisdata[OSind,])

table(irisdata$Species)
table(OSdata$Species)
```

*Undersampling* removes majority-class samples to match the minority-class size.
```{r}
USind=sample(majorind,majorn-minorn,replace=FALSE)
USdata<-irisdata[-USind,]

table(irisdata$Species)
table(USdata$Species)
```

*Note.* Oversampling can increase the risk of overfitting (by repeating minority examples), while undersampling may discard useful information from the majority class.

### SMOTE (Synthetic Minority Oversampling Technique)

In imbalanced classification problems, simply duplicating minority-class examples (oversampling) can lead to overfitting, while discarding majority-class examples (undersampling) may waste information. The SMOTE provides an alternative by generating *synthetic minority* observations. These new samples are created by interpolating between existing minority-class observations and their nearest neighbors.

We illustrate SMOTE using the `smotefamily` package on an imbalanced version of the `iris` dataset, where setosa is the majority and versicolor is the minority class.
```{r}
# install.packages("smotefamily")
library(smotefamily)

# Convert to numeric labels for smotefamily
x <- irisdata[,1:4]
y <- as.numeric(irisdata$Species)  # setosa=1, versicolor=2

# Apply SMOTE
set.seed(1)
iris.SMOTE <- SMOTE(x, y, K=5, dup_size=2)

table(y)                       # original distribution
table(iris.SMOTE$data$class)   # after SMOTE
```

Here:
`dup_size = 2` roughly doubles the minority class by generating synthetic examples.

`K = 5` specifies the number of nearest neighbors used when interpolating new synthetic points.

The `table()` commands show how the class distribution becomes more balanced after SMOTE.


*Note*. SMOTE creates more diverse samples than random oversampling, but if the classes overlap significantly, synthetic points may fall into ambiguous regions. Careful use of SMOTE is recommended when decision boundaries are not well-separated.
