---
title: "Unsupervised Learning I - Clustering Analysis"
author: "Patrick PUN Chi Seng (NTU Sg)"
output:
  html_document:
    df_print: paged
  pdf_document: default
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyfoot[C]{\thepage}
- \fancypagestyle{plain}{\pagestyle{fancy}}
- \fancyhead[L]{Machine Learning in Finance}
- \fancyhead[R]{R Lab - Clustering Analysis}
---


```{r global_options, include=FALSE}
library(knitr)
knitr::opts_chunk$set(eval=TRUE)
```

## References

Chapter 12.5.3, [ISLR2] An Introduction to Statistical Learning - with Applications in R (2nd Edition). Free access to download the book: https://www.statlearning.com/

To see the help file of a function `funcname`, type `?funcname`.

## $K$-Means Clustering

The function `kmeans()` performs $K$-means clustering in `R`.  We begin with a simple simulated example in which there truly are two clusters in the data: the first 25 observations have a mean shift relative to the next 25 observations.

```{r chunk28}
set.seed(1)
x <- matrix(rnorm(50 * 2), ncol = 2)
x[1:25, 1] <- x[1:25, 1] + 3
x[1:25, 2] <- x[1:25, 2] - 4
```

We now perform $K$-means clustering with $K=2$.

```{r chunk29}
km.out <- kmeans(x, 2, nstart = 20)
```

The cluster assignments of the 50 observations are contained in  `km.out$cluster`.

```{r chunk30}
km.out$cluster
```

The $K$-means clustering perfectly separated the observations into two clusters even though we did not supply any group information to `kmeans()`. We can plot the data, with each observation colored according to its cluster assignment.

```{r chunk31}
#par(mfrow = c(1, 2))
plot(x, col = (km.out$cluster + 1),
    main = "K-Means Clustering Results with K = 2",
    xlab = "", ylab = "", pch = 20, cex = 2)
```

<!-- Here the observations can be easily plotted because they are two-dimensional. If there were more than two variables then we could instead perform PCA and plot the first two principal components score vectors. -->

In this example, we knew that there really were two clusters because we generated the data. However, for real data, in general we do not know the true number of clusters. We could instead have performed $K$-means clustering on this example with $K=3$.

```{r chunk32}
set.seed(2)
km.out <- kmeans(x, 3, nstart = 20)
km.out
plot(x, col = (km.out$cluster + 1),
    main = "K-Means Clustering Results with K = 3",
    xlab = "", ylab = "", pch = 20, cex = 2)
```

When $K=3$, $K$-means clustering  splits up the two clusters.

To run the `kmeans()` function in `R` with multiple initial cluster assignments, we use the `nstart` argument. If a value of `nstart` greater than one is used, then $K$-means clustering will be performed using multiple random assignments in Step 1 of K-means Clustering Algorithm, and the `kmeans()` function will report only the best results. Here we compare using `nstart = 1` to `nstart = 20`.

```{r chunk33}
set.seed(3)
km.out <- kmeans(x, 3, nstart = 1)
km.out$tot.withinss
km.out <- kmeans(x, 3, nstart = 20)
km.out$tot.withinss
```

Note that `km.out$tot.withinss` is the total within-cluster sum of squares, which  we seek to minimize by performing $K$-means clustering. The individual within-cluster sum-of-squares are contained in the vector `km.out$withinss`.

We *strongly* recommend always running $K$-means clustering with a large value of `nstart`, such as 20 or 50, since otherwise an undesirable local optimum may be obtained.

When performing $K$-means clustering, in addition to using multiple initial cluster assignments, it is also  important to set a random seed using the `set.seed()` function. This way, the initial cluster assignments in Step 1 can be replicated, and the $K$-means output will be fully reproducible.

## Hierarchical Clustering

The `hclust()` function implements  hierarchical clustering in `R`. In the following example we use the simulated data before to plot the hierarchical clustering dendrogram using complete, single, and average linkage clustering, with  Euclidean distance as the dissimilarity measure.

We begin by clustering observations using complete linkage. The `dist()` function is used to compute the $50 \times 50$ inter-observation Euclidean distance matrix.

```{r chunk34}
hc.complete <- hclust(dist(x), method = "complete")
```

We could just as easily perform hierarchical clustering with average or single linkage instead:

```{r chunk35}
hc.average <- hclust(dist(x), method = "average")
hc.single <- hclust(dist(x), method = "single")
```

We can now plot the dendrograms obtained using the usual `plot()` function. The numbers at the bottom of the plot identify each observation.

```{r chunk36}
par(mfrow = c(1, 3))
plot(hc.complete, main = "Complete Linkage",
    xlab = "", sub = "", cex = .9)
plot(hc.average, main = "Average Linkage",
    xlab = "", sub = "", cex = .9)
plot(hc.single, main = "Single Linkage",
    xlab = "", sub = "", cex = .9)
```


To determine the cluster labels for each observation associated with a given cut of the dendrogram, we can use the `cutree()` function:

```{r chunk37}
cutree(hc.complete, 2)
cutree(hc.average, 2)
cutree(hc.single, 2)
```

The second argument to `cutree()` is the number of clusters we wish to obtain.

For this data, complete and average linkage generally separate the observations into their correct groups. However, single linkage identifies one point as belonging to its own cluster. A more sensible answer is obtained when four clusters are selected, although there are still two singletons.

```{r chunk38}
cutree(hc.single, 4)
```

To scale the variables before performing hierarchical clustering of the observations, we use the `scale()` function:

```{r chunk39}
xsc <- scale(x)
plot(hclust(dist(xsc), method = "complete"),
    main = "Hierarchical Clustering with Scaled Features")
```

Correlation-based distance can be computed using the `as.dist()` function, which converts an arbitrary square symmetric matrix into a form that the `hclust()` function recognizes as a distance matrix. However, this only makes sense for data with at least three features since the absolute correlation between any two observations with measurements on two features is always 1. Hence, we will cluster a three-dimensional data set. This data set does not contain any true clusters.

```{r chunk40}
x <- matrix(rnorm(30 * 3), ncol = 3)
dd <- as.dist(sqrt((1 - cor(t(x)))/2))
plot(hclust(dd, method = "complete"),
    main = "Complete Linkage with Correlation-Based Distance",
    xlab = "", sub = "")
```

## NCI60 Data Example

Unsupervised techniques are often used in the analysis of genomic data. In particular, clustering and PCA are popular tools. We illustrate these techniques on the `NCI` cancer cell line microarray data, which consists of $6,830$ gene expression measurements on $64$ cancer cell lines.

```{r chunk41}
library(ISLR2)
nci.labs <- NCI60$labs
nci.data <- NCI60$data
```

Each cell line is labeled with a cancer type, given in `nci.labs`. We do not make use of the cancer types in performing clustering/PCA, as these are unsupervised techniques. But after performing clustering/PCA, we will check to see the extent to which these cancer types agree with the results of these unsupervised techniques.

The data has $64$ rows and $6,830$ columns.

```{r chunk42}
dim(nci.data)
```

We begin by examining the cancer types for the cell lines.

```{r chunk43}
nci.labs[1:4]
table(nci.labs)
```

### Clustering the Observations of the NCI60 Data

We now proceed to hierarchically cluster the cell lines in the `NCI` data, with the goal of finding out whether or not the observations cluster into distinct types of cancer. To begin, we standardize the variables to have mean zero and standard deviation one. As mentioned earlier, this step is optional and should be performed only if we want each gene to be on the same *scale*.

```{r chunk50}
sd.data <- scale(nci.data)
```

We now perform hierarchical clustering of the observations using complete, single, and average linkage. Euclidean distance is used as the dissimilarity measure.

```{r chunk51}
par(mfrow = c(1, 3))
data.dist <- dist(sd.data)
plot(hclust(data.dist), xlab = "", sub = "", ylab = "",
    labels = nci.labs, main = "Complete Linkage")
plot(hclust(data.dist, method = "average"),
    labels = nci.labs, main = "Average Linkage",
    xlab = "", sub = "", ylab = "")
plot(hclust(data.dist, method = "single"),
    labels = nci.labs,  main = "Single Linkage",
    xlab = "", sub = "", ylab = "")
```

We see that the choice of linkage certainly does affect the results obtained. Typically, single linkage will tend to yield *trailing* clusters: very large clusters onto which individual observations attach  one-by-one. On the other hand, complete and average linkage tend to yield more balanced, attractive clusters. For this reason, complete and average linkage are generally preferred to single linkage. Clearly cell lines within a single cancer type do tend to cluster together, although the clustering is not perfect. We will use complete linkage hierarchical clustering for the analysis that follows.

We can cut the dendrogram at the height that will yield a particular number of clusters, say four:

```{r chunk52}
hc.out <- hclust(dist(sd.data))
hc.clusters <- cutree(hc.out, 4)
table(hc.clusters, nci.labs)
```

There are some clear patterns. All the leukemia cell lines fall in cluster $3$, while the breast cancer cell lines are spread out over three different clusters. We can plot the cut on the dendrogram that produces these four clusters:

```{r chunk53}
par(mfrow = c(1, 1))
plot(hc.out, labels = nci.labs)
abline(h = 139, col = "red")
```

The `abline()` function draws a straight line on top of any existing plot in `R`. The argument `h = 139` plots a horizontal line at height $139$ on the dendrogram; this is the height that results in four distinct clusters. It is easy to verify that the resulting clusters are the same as the ones we obtained using `cutree(hc.out, 4)`.

 
Printing the output of `hclust` gives a useful brief summary of the object:

```{r chunk54}
hc.out
```

We claimed earlier that $K$-means clustering and hierarchical clustering with the dendrogram cut to obtain the same number of clusters can yield very different results. How do these `NCI` hierarchical clustering results compare to what we get if we perform $K$-means clustering with $K=4$?

```{r chunk55}
set.seed(2)
km.out <- kmeans(sd.data, 4, nstart = 20)
km.clusters <- km.out$cluster
table(km.clusters, hc.clusters)
```

We see that the four clusters obtained using hierarchical clustering and $K$-means clustering  are somewhat different. Cluster $4$ in $K$-means clustering is identical to cluster $3$ in hierarchical clustering. However, the other clusters differ: for instance, cluster $2$ in $K$-means clustering contains a portion of the observations assigned to cluster 1 by hierarchical clustering, as well as all of the observations assigned to cluster $2$ by hierarchical clustering.